NorduGrid Storage -- Overview and Design of a Service-Oriented Storage System

### 2009-02-05 doc/storage_system/ARCStorage_src/Storagepaper-llncs.tex
Euro-parified storage paper

%\documentclass[10pts, twocolumn]{article}
%\documentclass[twocolumn]{IEEEtran}
\documentclass[final]{ieee}
\usepackage { graphicx}
\usepackage{wrapfig}

\hyphenation{librar-ian librar-ians shep-herd bar-tender gate-way name-space}

\begin{document}
\pagestyle{headings} 
\title{NorduGrid Storage -- Overview and Design of a Service-Oriented Storage System}
%% \title{Service-Oriented ARC Storage; \\An Overview and Design}
%% \author{  Jon K. Nilsen$^{2,3}$ \and Salman Toor$^{1}$
%%   \and Zsombor Nagy$^{5}$ \and Bjarte Mohn$^{4}$  \\ $^{1}$Scientific
%%   Computing Division, IT Department, Uppsala University, Sweden
%%   \\ $^{2}$Center for Information Technology, University of Oslo, Norway
%%   \\ $^{3}$Department of Physics,  University of Oslo, Norway
%%   \\  $^{4}$Department of Physics and Astronomy, Uppsala University,
%%   Sweden \\ $^{5}$National Information Infrastructure Development
%%   Institute, Hungary}
\author{  Jon K. Nilsen$^{\dagger,\ddagger}$ \and Salman Toor$^{*}$
  \and Zsombor Nagy$^{\mathparagraph}$ \and Bjarte Mohn$^{\mathsection}$ 
 \\ $^{\dagger}$University of Oslo, Center for Information Technology,\\
P.~O.~Box~1059, Blindern, N-0316~Oslo, Norway
  \\ $^{\ddagger}$University of Oslo, Dept. of Physics, P.~O.~Box~1048,
Blindern, N-0316~Oslo, Norway
\\$^{*}$Dept. of Information Technology, Div. of Scientific Computing, Uppsala
University,\\ Box~337, SE-75105~Uppsala, Sweden\\
$^{\mathparagraph}$Institute of National Information and Infrastructure
Development\\ NIIF/HUNGARNET, Victor~Hugo~18-22, H-1132~Budapest, Hungary 
  \\  $^{\mathsection}$Dept. of Physics and Astronomy, Div. of Nuclear and
Particle Physics,\\ Uppsala University, Box~535, SE-75121~Uppsala, Sweden \\ \\
j.k.nilsen@usit.uio.no \hspace{5 mm}
  salman.toor@it.uu.se \hspace{5 mm} zsombor@niif.hu \\  \hspace{7 mm}
  bjarte.mohn@fysast.uu.se}

  % \\ \small{salman.toor \and bjarte.mohn  }}

@@

The challenge of building a reliable, self-healing, fault-tolerant,
consistent data management system at the web-scale is an interesting
task. Making the system work in a heterogeneous, distributed
environment like the Grid is even more interesting. An increasing
number of applications demand not only increased CPU power, but also
vast amounts of storage space. The required storage space is not only
restricted to the duration in which the application runs; the data
should often be available for years afterwards, in certain cases even
for decades. Nowadays, we can easily find single Grid jobs which produce
gigabytes or even terabytes of data, ramping up the requirements of
storage systems to the petabyte-scale and beyond. Hard drive capacity is
still increasing and storage raids can handle failing hard drives through 
data replication. However, even a designated storage site cannot guarantee 
24/7 availability or long-term security against catastrophic losses of data. 
The need for a distributed, self-healing storage system is evident. To
make the storage system useable to Grid users, the system must provide
reliable and secure file transfer protocols, a cataloging system and
secure storage of the data. Several projects and designs have emerged
to address such challenges~\cite{Hoschek00datamanagement,DengWang}.

In the advent of the next generation of the Advanced Resource
Connector (ARC) Grid middleware~\cite{arc} (new release due during
fall 2009), we present the new NorduGrid storage~\cite{ARCStoragedesigndoc}. This distributed storage system is
designed to provide an easy to use, flexible and scalable system that
can offer native storage and at the same time provide access to
third-party solutions like dCache~\cite{dCache,DSSWithdCache} by using
the same, uniform interface. Being part of ARC, the storage system is
based on a service-oriented architecture, in which each major
component of the system runs as a separate service within the ARC
Hosting Environment Daemon (HED)~\cite{HEDdesigndoc}. The HED service
container gives the capability of flexible replacement of the
components as well as the possibility to introduce modifications in
the future. %% The HED service container gives
%% the capability of flexible replacement of the components as well as
%% the future modification/enhancement of the system.

This paper is organized as follows. After describing
related work in Section \ref{Related Work}, we give a bird's eye
view of the next-generation Advanced Resource Connector (ARC) in
Section \ref{The Advance Resource Connector}. An overview of the NorduGrid storage is given in Section \ref{The NorduGrid Storage}, while
the architecture of the storage system is elaborated in Section
\ref{Architecture of the NorduGrid Storage}. In Section \ref{Testing and
  Discussion} we give some early, proof-of-concept results, before 
concluding in Section \ref{Conclusion and Future Work}.

\section{Related Work}
\label{Related Work}
\textbf{dCache:} dCache is a storage system which combines heterogeneous storage
elements to collect several hundreds of terabytes in a single
namespace. It additionally supports standard and native
protocols like gridftp, srm, dcap and gsidcap. dCache is
a joint effort between Deutsches Elektronen Synchrotron (DESY) \cite{dCachesite}, Fermi National Accelerator Laboratory \cite{FermiLab}, the  Nordic Data
Grid Facility (NDGF) \cite{NDGF}  and several other collaborators. %% The services or
%% components of dCache are known as cells. The standard deployment of
%% dCache consist of  storage pools; used to store file, optionally be
%% connected with HSM\footnote{Hierarchical Storage Manager, used for the
%%   tape storage}. Namespace manager; used PNFS \footnote{Perfectly
%%   Normal FileSystem; A network file system, designed in Desy} as a
%% filesystem  for storing matadata. And the number of protocol
%% doors. One of the biggest deployment of dCache is the distributed
%% infrastructure of  NDGF.\\ 
dCache has proven to be a very stable and scalable solution. However,
it is relatively difficult to deploy and integrate with new applications. 
The NorduGrid storage, being a light-weight and flexible
storage solution, aims more towards new user groups less familiar with
Grid solutions. 
\newline
\textbf{BigTable:} Bigtable is a distributed storage system managing structured data on
the petabyte-scale \cite{Bigtable}. It is currently used within
Google in projects such as web indexing, Google Earth and Google
Financing. %% Bigtable is mainly built on top of three pieces of Google's
%% infrastructure: The Google Filesystem \cite{GFS} a distributed,
%% replicating file system, the Google SSTable, providing
%% a persistent, ordered immutable map and Chubby \cite{Chubby}, a highly
%% available and persistent distributed lock service. The system uses a
%% three-level B$^+$-tree to store \textit{tablets}. The location of the
%% first level is stored in Chubby, making Chubby a vitaly important part
%% of the design. To ensure a persistent solution, Chubby takes advantage
%% of the Paxos algorithm \cite{Paxos,PaxosLive}.\\
Bigtable has several interesting features, one of which is the
distributed lock service, Chubby \cite{Chubby}. Chubby is a high-availability, distributed
locking service sitting on top of the B$^+$-tree architecture of
Bigtable. Chubby has several features similar to our A-Hash service (see Section \ref{Hash Algorithm}),
among which the use of the Paxos algorithm \cite{Paxos,PaxosLive}, is
the most striking. A major caveat for the Grid community is that
Bigtable is neither free, open-source nor available to the public.
\newline
\textbf{Storage Resource Broker:}  Based on the client-server model,
the Storage Resource Broker (SRB) \cite{earlySRB,SRB} provides a flexible data grid management
system. It allows  uniform access to heterogeneous
storage resources over a wide area network. %% SRB also manages the
%% namespaces used to identify files, storage resources, users, metadata,
%% and access controls independently of the physical storage systems. It
%% uses centralized matadata catalog (MCat). Also supports several file
%% systems of UNIX, Windows, Linux, Mac OS X, HPSS and large scale
%% storage systems like Castor and dCache.\\
Its functionality, with a uniform namespace for several Data Grid
Managers and file systems, is quite similar to the functionality
offered by our Gateway service (see Section \ref{Gateways}). However, being built as a middleware on top of
other major storage solutions, SRB does not offer its own storage solution.
\newline
\textbf{Scalla:} Scalla is a widely used software suite consisting of an
xrootd server for data access, and an olbd
server for building scalable xrootd clusters \cite{Scalla}. %% The xrootd
%% server consists of several plug-in based components, of which the most
%% important are; the xrd component, responsible for network, threading,
%% data buffer and protocol management; the xroot protocol component,
%% providing the specialized and increasingly popular xroot protocol; and
%% components for logical and physical file systems. The olbd server is
%% used for setting up storage clusters in a B-64 tree with a
%% redirector on top appointing one supervisor for every 64 data servers
%% in the lower layers of the tree.
Originally developed for use with the physics analysis tool root \cite{root},
xrootd offers data access both through the specialized xroot protocol
and through other third-party protocols. The combination of the xrootd
and olbd components offers a cluster storage designed for low latency,
high bandwidth environments. In contrast, the NorduGrid storage is optimized
for high latency and is more suitable for the Grid environment.

@@

To deliver the non-trivial quality of services required by the Grid,
there are a number of services running inside the HED. For example, Grid
job execution and management is handled by the A-REX service 
\cite{AREXdesigndoc}, policy
decisions are taken by the Charon service, the ISIS service is
responsible for information indexing, batch job submission is
handled by the Sched service, etc. In the later sections our discussion
will focus on the architecture and the design of another major set
of services, i.e., the NorduGrid storage.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{arc1-storage-services-with-gateway.pdf}
  \end{center}
  \caption{Schematic of the NorduGrid storage architecture. The figure shows
    the main services of the NorduGrid storage and the communication channels. The B stands for Bartender, the S for Shepherd, the L for Librarian, the G for Gateway and the A-H stands for A-Hash. The straight lines denote the communication between services.}
  \label{fig:arc1_storage_service}
\end{figure}

Being based on a service-oriented architecture, the NorduGrid storage consists of a set of services as shown in Fig.
\ref{fig:arc1_storage_service}. The services are as follows: The Bartender (B)
provides the high-level interface to the user; the Librarian (L)
handles the entire storage namespace, using the A-Hash (A-H) as a metadatabase; the
Gateway (G) provides access to third-party storage systems; and
the Shepherd (S) is the frontend for the physical storage node. 
See Section \ref{Architecture of the NorduGrid Storage} for a detailed
discussion of the different services.   
The services communicate with each other through the
Message Chain Components in HED. The communication channels are
depicted by straight lines in Fig. \ref{fig:arc1_storage_service}.

The system supports file transfer through several transfer protocols,
with client side tools that hide various technical
details such as protocol specification, port numbers
and so on. To provide fault-tolerance, the system implements
automatic file replication, where
the replicas of a file\footnote{In this paper, a file
denotes an entry in the global namespace, while replica denotes the
physical file stored on a storage node.} are always stored on separate
storage nodes\footnote{A
  storage node is a server with a Shepherd, a Shepherd backend
and some storage service.}. To ensure a resilient, self-healing system, the Shepherd
regularly sends heartbeats to one of the Librarians. If the Shepherd fails
to send a heartbeat, one of the Librarians will mark its replicas as offline, thus initiating
re-replication of the replicas between the other Shepherds.

In the default implementation, the services will provide a
full-featured and consistent data storage system using files as an
atomic unit. The scope of the NorduGrid storage in the forseeable
future is restricted to providing support for file-based data services.
%% there are currently no plans for intrinsic support for database
%% access.

The NorduGrid storage supports third-party storage services in two
ways. Using the Gateway service, files already stored in some
third-party storage can be accessed using the provided client tools. To make use of third-party services for storing new
files, with the replication and fault-tolerance offered by the NorduGrid storage, third-party storage elements can also be used as backends for the
Shepherd service.

External grid middleware components can
access the storage system using ARC data service
interfaces directly. ARC will also provide interface
components that communicate via standard protocols such as SRM, which
will provide a single access point to the whole system.

@@

\subsection{Core Components} 
\subsubsection{Bartender}
\label{Bartender}
The Bartender service provides a high-level interface for the storage
system. Clients connect to the Bartender to create and remove
files, collections and mount-points using their Logical Names. The
Bartender communicates 
with the Librarian and Shepherd services to execute the clients'
requests. However, the actual file data does not go through the
Bartender, instead file
transfers are directly performed between the storage nodes and the
clients. There could be any number of independent Bartender services
running in the system, providing high-availability and
load-balancing. 
\subsubsection{Librarian}
\label{librarian}
The Librarian manages the hierarchy and metadata of
files, collections and mount points, as well as the health information of the Shepherd
services. In addition, the Librarian handles
the information about registered Shepherd services. The Librarian receives
heartbeat messages from the Shepherds and changes replica states automatically
if needed. The Librarian uses the A-Hash for consistently storing all metadata. This makes the
Librarian a stateless service, thus enabling the system to have any
number of independent
Librarian services, again providing high-availability in the
system.

\subsubsection{Shepherd}
\label{Shepherd}

The Shepherd services run as front-ends on storage nodes. A Shepherd
service reports to a Librarian about the 
node's health state in terms of replicas. While the Bartender initiates
file transfers, the actual transfers go directly between the storage node
and the clients.

@@

\subsubsection{A-Hash}
\label{Hash Algorithm}

A-Hash is a hash table for consistently storing data in property-value
pairs. All metadata about files, collections, mount point, Shepherd's
health status, and so forth, is stored in the A-Hash. As the A-Hash is
the service storing the entire state of the storage system, it is
absolutely crucial for the sttorage system that the A-Hash is
consistent. The distribution and replication of this service is
therefore both necessary and challenging.

\begin{figure}
\centering
\includegraphics[width=1.0\columnwidth]{arc1-storage-downloading.pdf}
\caption{Schematic showing the scenario of file download from the NorduGrid storage.}
\label{arc1-storage-downloading}
\end{figure}

\subsection{Features}
\subsubsection{Heartbeat monitoring}
\label{Heartbeat and Replication}
In the proposed architecture, each Shepherd periodically sends
heartbeats to a Librarian with information about replicas
whose state changed since the last heartbeat, and the corresponding
GUIDs. These heartbeats are then stored in the A-Hash, making them
visible to all the Librarians in the system. If any of the Librarians
notices that a Shepherd is late with its heartbeat, it will mark
all the replicas in that Shepherd as offline.

\subsubsection{Replication}
\label{Replication}
Shepherds periodically ask the Librarian if the file of the replica stored on
its storage node has enough replicas. If the file does not have
enough replicas, the Shepherd informs the Bartender and the Bartender initiates
a put request that returns a Transfer URL to the Shepherd. The
Shepherd finally uploads the new replica. The Shepherd which gets the new replica notifies the
Librarian that the replica is alive. The Librarian then adds this to
the corresponding file.

\subsubsection{Deletion}
\label{Deletion}
In a replicating storage system, there are several possible solutions
for file deletion, since both the replicas and the metadata need to be
removed. In our solution, we use the process of \textit{lazy
  deletion} \cite{LazyDeletion}. When a client requests that a file
should be deleted (and the user has the proper permissions) the
Bartender instructs the Librarian to remove the file's GUID from the
A-Hash. When the Shepherd, periodically checking all its replicas,
discovers that a replica has no file (and hence, no Location), it will
automatically delete the replica.

\subsubsection{Fault tolerance}
\label{Fault tolerance}

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.2 for LaTeX2e
%

\documentclass{llncs}
\usepackage{graphicx}
\usepackage{wrapfig}
\hyphenation{librar-ian librar-ians shep-herd bar-tender gate-way name-space}
%
\begin{document}
%

\pagestyle{headings} 
\mainmatter              % start of the contributions

\title{NorduGrid Storage -- Overview and Design of a Service-Oriented Storage System}

%\title{Service-Oriented ARC Storage;  An Overview and Design}

% \titlerunning{LAP-GJMF Integration}  % abbreviated title (for running head)
%                                         also used for the TOC unless
%                                         \toctitle is used
%
\author{Jon K. Nilsen\inst{1} \and  Salman Toor\inst{2} \and
Zsombor Nagy\inst{3} \and Bjarte Mohn\inst{4} }
%
%\authorrunning{} % abbreviated author list (for running head)
%
%%%% modified list of authors for the TOC (add the affiliations)
\tocauthor{Jon K. Nilsen (Oslo University),
 Salman Toor (Uppsala University),
 Zsombor Nagy (NIIF),
 Bjarte Mohn (Particle Physics Uppsala University)}

\institute{ University of Oslo, Dept. of Physics,  P. O. Box 1048, Blindern, N-0316 Oslo, Norway\\University of Oslo, Center for Information Technology, P. O. Box 1059, Blindern, N-0316 Oslo, Norway\\
\email{j.k.nilsen@usit.uio.no}
\and
Dept. Information Technology, Div. of Scientific Computing Uppsala University, Box 256, SE-751 05 Uppsala, Sweden\\
\email{salman.toor@it.uu.se http://www.uu.se}
\and
 Institute of National Information and Infrastructure Development NIIF/HUNGARNET, Victor Hugo 18-22, H-1132 Budapest, Hungary
\email{zsombor@niif.hu}
\and
Dept. of Physics and Astronomy, Div. of Nuclear and Particle Physics, Uppsala University, Box 535, SE-75121 Uppsala, Sweden\\
\email{bjarte.mohn@fysast.uu.se}
}

@@

The challenge of building a reliable, self-healing, fault-tolerant,
consistent data management system at the web-scale is an interesting
task. Making the system work in a heterogeneous, distributed
environment like the Grid is even more interesting. An increasing
number of applications demand not only increased CPU power, but also
vast amounts of storage space. The required storage space is not only
restricted to the duration in which the application runs; the data
should often be available for years afterwards, in certain cases even
for decades. Nowadays, we can easily find single Grid jobs which produce
gigabytes or even terabytes of data, ramping up the requirements of
storage systems to the petabyte-scale and beyond.
%jon
%which requires the need
%salman
 %Hard drive capacity is 
 %still increasing and storage raids can handle failing hard drives through 
%data replication. However, even a designated storage site cannot guarantee 
%24/7 availability or long-term security against catastrophic losses of data. The need 
%for a distributed, self-healing storage system is evident. 
To make the storage system useable to Grid users, the system must provide
reliable and secure file transfer protocols, a cataloging system and
secure storage of the data. Several projects and designs have emerged
to address such challenges~\cite{Hoschek00datamanagement,DengWang}.

In the advent of the next generation of the Advanced Resource
Connector (ARC) Grid middleware~\cite{arc} (new release due during
fall 2009), we present the new NorduGrid storage~\cite{ARCStoragedesigndoc}. This distributed storage system is
designed to provide an easy to use, flexible and scalable system that
can offer native storage and at the same time provide access to
third-party solutions
%salman
% like dCache~\cite{dCache,DSSWithdCache} 
by using the same, uniform interface.

%salman
%Being part of ARC, the storage system is
%based on a service-oriented architecture, in which each major
%component of the system runs as a separate service within the ARC
%Hosting Environment Daemon (HED)~\cite{HEDdesigndoc}. The HED service
%container gives the capability of flexible replacement of the
%components as well as the possibility to introduce modifications in
%the future.

%% The HED service container gives
%% the capability of flexible replacement of the components as well as
%% the future modification/enhancement of the system.

This paper is organized as follows. After describing
general Grid storage systems in \ref{Characteristics of Grid Storage}
and related work in Section \ref{Related Work}, we give a bird's eye
view of the next-generation Advanced Resource Connector (ARC) in
Section \ref{The Advance Resource Connector}. An overview of the NorduGrid storage is given in Section \ref{The NorduGrid Storage}, while
the architecture of the storage system is elaborated in Section
\ref{Architecture of the NorduGrid Storage}. In Section \ref{Testing and
  Discussion} we give some early, proof-of-concept results, before 
concluding in Section \ref{Conclusion and Future Work}.

%jon
\section{Characteristics of Grid Storage}
\label{Characteristics of Grid Storage}

A Grid storage system provides its clients with access to data
stored at remote storage systems. A traditional architecture (see
e.g. \cite{datagrid,egeeddm}) typically
consists of an {\it indexing service}, indexing files from storage
resources, a {\it replication service} for managing replica locations,
and a (often centralized) {\it metadata catalog} imposing a global
namespace on top of the resources. Data access can be handled either
through a file transfer service or directly through the storage
resource by quering the metadata catalog.

While this architecture has strongly improved the accessability of
physically distributed data, common Grid storage solutions typically have
some limitations: The metadata catalog, which is in most cases
centralized can quickly become a bottleneck. In typical Grid storage
systems, the storage resource is unaware of the state of its files,
making consistency guarantees hard to give. Traditional Grid
solutions often take for granted that the resource are dedicated for
the Grid, and in some cases even require specific operating systems,
thus restricting the amount of available hardware resources\footnote{It is often hard to convince a
system administrator to install a specific operating system to be able to share
her/his hardware resources in a Grid.}. In addition, lack of
standardization often makes interoperability between storage systems
non-trivial.
The NorduGrid Storage has as a goal to meet these limitations to
provide a truly distributed, self-healing, flexible Grid storage
solution, with a replicated metadata catalog, state aware storage
resources and an operating system agnostic implementation.
%/jon

\section{Related Work}
\label{Related Work}
%jon
Many Grid projects have developed their own, aplication specific
solutions. However, some general Grid storage solutions are widely
adopted:

dCache~\cite{dCache} is a service oriented storage system which combines heterogeneous storage
elements to collect several hundreds of terabytes in a single
namespace. Originally designed to work on a local area network,
dCache has proven to be usefull also in a Grid environment, with the
NDGF (Nordic Data Grid Facillity) dCache
installation~\cite{DSSWithdCache} as the largest example. There, the
core components, such as the metadata catalog, indexing service and
protocol doors are run in a centralized manner, while the storage
pools are distributed. The
NorduGrid storage, designed to have multiple instances of all
services running in a Grid environment, will not need a centralized
set of core services. Additionally, dCache is relatively difficult to
deploy and integrate with new applications. Being a more light-weight and flexible
storage solution, the NorduGrid storage aims more towards new user
groups less familiar with Grid solutions.

%% dCache~\cite{dCache} is a storage system which combines heterogeneous storage
%% elements to collect several hundreds of terabytes in a single
%% namespace. %% It additionally supports standard and native
%% protocols like gridftp, srm, dcap and gsidcap.
%jon
%dCache is
%a joint effort between Deutsches Elektronen Synchrotron (DESY) \cite{dCachesite}, Fermi National Accelerator Laboratory \cite{FermiLab}, the  Nordic Data
%Grid Facility (NDGF) \cite{NDGF}  and several other collaborators. %% The services or
%% components of dCache are known as cells. The standard deployment of
%% dCache consist of  storage pools; used to store file, optionally be
%% connected with HSM\footnote{Hierarchical Storage Manager, used for the
%%   tape storage}. Namespace manager; used PNFS \footnote{Perfectly
%%   Normal FileSystem; A network file system, designed in Desy} as a
%% filesystem  for storing matadata. And the number of protocol
%% doors. One of the biggest deployment of dCache is the distributed
%% infrastructure of  NDGF.\\ 
%% dCache has proven to be a fairly stable and scalable solution. However,
%% it is relatively difficult to deploy and integrate with new applications. 
%% The NorduGrid storage, being a light-weight and flexible
%% storage solution, aims more towards new user groups less familiar with
%% Grid solutions. 
%jon
%% \textbf{BigTable:} Bigtable is a distributed storage system managing structured data on
%% the petabyte-scale \cite{Bigtable}. It is currently used within
%% Google in projects such as web indexing, Google Earth and Google
%% Financing. %% Bigtable is mainly built on top of three pieces of Google's
%% %% infrastructure: The Google Filesystem \cite{GFS} a distributed,
%% %% replicating file system, the Google SSTable, providing
%% %% a persistent, ordered immutable map and Chubby \cite{Chubby}, a highly
%% %% available and persistent distributed lock service. The system uses a
%% %% three-level B$^+$-tree to store \textit{tablets}. The location of the
%% %% first level is stored in Chubby, making Chubby a vitaly important part
%% %% of the design. To ensure a persistent solution, Chubby takes advantage
%% %% of the Paxos algorithm \cite{Paxos,PaxosLive}.\\
%% Bigtable has several interesting features, one of which is the
%% distributed lock service, Chubby \cite{Chubby}. Chubby is a high-availability, distributed
%% locking service sitting on top of the B$^+$-tree architecture of
%% Bigtable. Chubby has several features similar to our A-Hash service, %(see Section \ref{Hash Algorithm}),
%% among which the use of the Paxos algorithm \cite{Paxos,PaxosLive}, is
%% the most striking. A major caveat for the Grid community is that
%% Bigtable is neither free, open-source nor available to the public.
%jon

XTtreemFS~\cite{xtreemfs} is an object-based
file system specifically designed for the Grid environment of the
operating system XTreemOS. Being object-based, XTreemFS have separated
the metadata of files and the file content, or {\it objects} in the
same fashion as the NorduGrid storage. Being
a tightly integrated part of an operating system enables XTreemFS to
implement functionality very similar to that of local file
systems while being a distributed Grid storage. However, this also enforces
a specific operating system, requiring dedicated hardware resources.
%/jon

Based on the client-server model,
the Storage Resource Broker (SRB) \cite{earlySRB,SRB} provides a flexible data grid management
system. It allows  uniform access to heterogeneous
storage resources over a wide area network. %% SRB also manages the
%% namespaces used to identify files, storage resources, users, metadata,
%% and access controls independently of the physical storage systems. It
%% uses centralized matadata catalog (MCat). Also supports several file
%% systems of UNIX, Windows, Linux, Mac OS X, HPSS and large scale
%% storage systems like Castor and dCache.\\
Its functionality, with a uniform namespace for several Data Grid
Managers and file systems, is quite similar to the functionality
offered by our Gateway service (see Section \ref{Features}). However, being built as a middleware on top of
other major storage solutions, SRB does not offer its own storage solution.

Scalla is a widely used software suite consisting of an
xrootd server for data access, and an olbd
server for building scalable xrootd clusters \cite{Scalla}. %% The xrootd
%% server consists of several plug-in based components, of which the most
%% important are; the xrd component, responsible for network, threading,
%% data buffer and protocol management; the xroot protocol component,
%% providing the specialized and increasingly popular xroot protocol; and
%% components for logical and physical file systems. The olbd server is
%% used for setting up storage clusters in a B-64 tree with a
%% redirector on top appointing one supervisor for every 64 data servers
%% in the lower layers of the tree.
Originally developed for use with the physics analysis tool root \cite{root},
xrootd offers data access both through the specialized xroot protocol
and through other third-party protocols. The combination of the xrootd
and olbd components offers a cluster storage designed for low latency,
high bandwidth environments. In contrast, the NorduGrid storage is optimized
for high latency and is more suitable for the Grid environment.

@@

%salman
%To deliver the non-trivial quality of services required by the Grid,
%there are a number of services running inside the HED. For example, Grid
%job execution and management is handled by the A-REX service 
%\cite{AREXdesigndoc}, policy
%decisions are taken by the Charon service, the ISIS service is
%responsible for information indexing, batch job submission is
%handled by the Sched service, etc. In the later sections our discussion
%will focus on the architecture and the design of another major set
%of services, i.e., the NorduGrid storage.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.35\columnwidth]{arc1-storage-services-with-gateway.pdf}
  \end{center}
  \caption{Schematic of the NorduGrid storage architecture. The figure shows
    the main services of the NorduGrid storage and the communication channels. The B stands for Bartender, the S for Shepherd, the L for Librarian, the G for Gateway and the A-H stands for A-Hash. The straight lines denote the communication between services.}
  \label{fig:arc1_storage_service}
\end{figure}

Being based on a service-oriented architecture, The NorduGrid storage consists of a set of services as shown in Fig.
\ref{fig:arc1_storage_service}. The services are as follows: The Bartender (B)
provides the high-level interface to the user; the Librarian (L)
handles the entire storage namespace, using the A-Hash (A-H) as a metadatabase; the
Gateway (G) provides access to third-party storage systems; and
the Shepherd (S) is the frontend for the physical storage node. 
See Section \ref{Architecture of the NorduGrid Storage} for a detailed
discussion of the different services.   
The services communicate with each other through the
Message Chain Components in HED. The communication channels are
depicted by straight lines in Fig. \ref{fig:arc1_storage_service}.

%salman
%The system supports file transfer through several transfer protocols,
%with client side tools that hide various technical
%details such as protocol specification, port numbers
%and so on. To provide fault-tolerance, the system implements
%automatic file replication, where
%the replicas of a file\footnote{In this paper, a file
%denotes an entry in the global namespace, while replica denotes the
%physical file stored on a storage node.} are always stored on separate
%storage nodes\footnote{A
%  storage node is a server with a Shepherd, a Shepherd backend
%and some storage service.}. To ensure a resilient, self-healing system, the Shepherd
%regularly sends heartbeats to one of the Librarians. If the Shepherd fails
%to send a heartbeat, one of the Librarians will mark its replicas as offline, thus initiating
%re-replication of the replicas between the other Shepherds.

%In the default implementation, the services will provide a
%full-featured and consistent data storage system using files as an
%atomic unit. The scope of the NorduGrid storage in the forseeable
%future is restricted to providing support for file-based data services.

%% there are currently no plans for intrinsic support for database
%% access.

%salman
%The NorduGrid storage supports third-party storage services in two
%ways. Using the Gateway service, files already stored in some
%third-party storage can be accessed using the provided client tools. To make use of third-party services for storing new
%files, with the replication and fault-tolerance offered by the NorduGrid storage, third-party storage elements can also be used as backends for the
%Shepherd service.

%External grid middleware components can
%access the storage system using ARC data service
%interfaces directly. ARC will also provide interface
%components that communicate via standard protocols such as SRM, which
%will provide a single access point to the whole system.

@@

\subsection{Core Components} 
%jon add
The core components (Fig. \ref{fig:arc1_storage_service}) are
described as follows:
\begin{itemize}
%\subsubsection{Bartender}
%\label{Bartender}

\item The \textbf{Bartender} %The Bartender 
provides a high-level interface for the storage
system. Clients connect to the Bartender to create and remove
files, collections and mount-points using their Logical Names. The
Bartender communicates 
with the Librarian and Shepherd services to execute the clients'
requests. However, the actual file data does not go through the
Bartender, instead file
transfers are directly performed between the storage nodes and the
clients. There could be any number of independent Bartender services
running in the system, providing high-availability and
load-balancing.

%\subsubsection{Librarian}
%\label{librarian}

\item  The {\bf Librarian} 
manages the hierarchy and metadata of
files, collections and mount points, as well as the health information of the Shepherd
services. In addition, the Librarian handles
the information about registered Shepherd services. The Librarian receives
heartbeat messages from the Shepherds and changes replica states automatically
if needed. The Librarian uses the A-Hash for consistently storing all metadata. This makes the
Librarian a stateless service, thus enabling the system to have any
number of independent
Librarian services, again providing high-availability in the
system.

%\subsubsection{Shepherd}
%\label{Shepherd}

\item The {\bf Shepherd} services 
runs as front-ends on storage nodes. A Shepherd
service reports to a Librarian about the 
node's health state in terms of replicas. While the Bartender initiates
file transfers, the actual transfers go directly between the storage node
and the clients.

@@

%\subsubsection{A-Hash}
%\label{Hash Algorithm}

\item The \textbf{A-Hash}  %A-Hash 
is a hash table for consistently storing data in property-value
pairs. All metadata about files, collections, mount point, Shepherd's
health status, and so forth, is stored in the A-Hash. As the A-Hash is
the service storing the entire state of the storage system, it is
absolutely crucial for the sttorage system that the A-Hash is
consistent. The distribution and replication of this service is
therefore both necessary and challenging.

%jon
\end{itemize}
%% \begin{figure}
%% \centering
%% \includegraphics[width=0.55\columnwidth]{arc1-storage-downloading.pdf}
%% \caption{Schematic showing the scenario of file download from the NorduGrid storage.}
%% \label{arc1-storage-downloading}
%% \end{figure}

\subsection{Features}
\label{Features}
\begin{itemize}

%\subsubsection{Heartbeat monitoring}
%\label{Heartbeat and Replication}
\item \textbf{Heartbeat monitoring:} In the proposed architecture, each Shepherd periodically sends
heartbeats to a Librarian with information about replicas
whose state changed since the last heartbeat, and the corresponding
GUIDs. These heartbeats are then stored in the A-Hash, making them
visible to all the Librarians in the system. If any of the Librarians
notices that a Shepherd is late with its heartbeat, it will mark
all the replicas in that Shepherd as offline.

%\subsubsection{Replication}
%\label{Replication}

\item \textbf{Replication:} Shepherds periodically ask the Librarian if the file of the replica stored on
its storage node has enough replicas. If the file does not have
enough replicas, the Shepherd informs the Bartender and the Bartender initiates
a put request that returns a Transfer URL to the Shepherd. The
Shepherd finally uploads the new replica. The Shepherd which gets the new replica notifies the
Librarian that the replica is alive. The Librarian then adds this to
the corresponding file.

%\subsubsection{Deletion}
%\label{Deletion}

%jon
\item \textbf{Deletion:} In a replicating storage system, there are several possible solutions
for file deletion, since both the replicas and the metadata need to be
removed. In our solution, we use the process of \textit{lazy
  deletion} \cite{LazyDeletion}. When a client requests that a file
should be deleted (and the user has the proper permissions) the
Bartender instructs the Librarian to remove the file's GUID from the
A-Hash. When the Shepherd, periodically checking all its replicas,
discovers that a replica has no file (and hence, no Location), it will
automatically delete the replica.

%\subsubsection{Fault tolerance}
%\label{Fault tolerance}

\item \textbf{Fault tolerance:} Due to the unpredictable nature of the Grid environment, it is essential to
have some degree of  automatic recovery system in case of unexpected
failures. Fault tolerant behavior is required both at the level of
metadata and on the level of physical storage. While the work on fault-tolerant metadata is still in
progress, the following two
scenarios will explain the currently available recovery mechanism for
physical storage: 
\begin{itemize}
\item In the case of a file having an invalid checksum, the Shepherd
immediately informs the Librarian and the Librarian changes the state of the given replica to
`\textbf{invalid}'. To recover its replica, the Shepherd contacts a
Bartender and asks for another replica of the file. The Bartender chooses a valid
replica, initiates a file transfer from a Shepherd having the replica,
and returns the TURL to the Shepherd with the invalid replica. When the
Shepherd has received the replica and compared the checksum, it notifies the
Librarian that the replica is alive again.  
\item In the case of a Shepherd going offline, a Librarian will,
  as mentioned earlier, notice the lack of heartbeats and invalidate all the
  replicas, initiating new replication for all the files stored in
  this storage node. However, if the Shepherd
  again comes online, there will evidently be more replicas than
  needed. The first Shepherd to notice this will set its replica's
  state to `\textbf{thirdwheel}', i.e. obsolete. At the next occasion,
  the Shepherd will remove the replica, if and only if it has the only
  `\textbf{thirdwheel}' replica of this file. If there are more
  replicas with this state, all replicas will be set back to
  `\textbf{alive}' and the process is repeated. This scenario will be discussed further in Section
  \ref{Testing and Discussion}.
\end{itemize}

%\subsubsection{Client tools }
%\label{Client extension using FUSE module}
